---
title: "BART Passenger Heatmap"
author: "Justin Perona"
date: "2019-03-20"

# code courtesy of R Markdown: The Definitive Guide by Xie, Allaire, and Grolemund
# https://bookdown.org/yihui/rmarkdown/html-document.html
runtime: shiny
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: false
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

In Winter Quarter 2018, I took ECI 254 (Exploring Data from Built Environment Using R) with Professor Deb Niemeier.
My final project for that class was to create a heatmap of passengers that traveled through the San Francisco Bay Area Rapid Transit (BART) system.

### Previous Work and Motivation

In Spring Quarter 2018, I took MAE 253 (Network Theory) with Professor Raissa D'Souza.
My final project for that class involved analyzing the network formed by the BART system with a team.
One of the papers I read for that project, [*Weighted complex network analysis of travel routes on the Singapore public transportation system*](https://doi.org/10.1016/j.physa.2010.08.015.) by Soh et al., argued that a dynamic (weighted) network analysis of a metro rail network like BART could give different insights into a network than a topological (unweighted) analysis.
In order to analyze the BART network, I wanted to create visualizations of the dynamic BART network at different times and days.
To do so, I ended up writing a Python script that would parse the BART hourly passenger origin-destination data [available on their website](https://www.bart.gov/about/reports/ridership) into a format that could be easily plotted in an external tool.
This allowed me to create the visualizations my team needed for the project.
That Python script and documentation on how to use it is available publicly on GitHub, in my repository [`jlperona/bart-hourly-dataset-parser`](https://github.com/jlperona/bart-hourly-dataset-parser).

After the class, I had a conversation with one of my teammates about how to take this work further.
I thought about creating a heatmap of the BART network using the same data I worked with.
The heatmap would make it easy to see at a glance where the majority of the passengers were travelling.
My teammate's suggestion was to use a visualization library like [D3.js](https://d3js.org/).
However, I didn't have experience with JavaScript, so I ended up shelving the idea.

When the final project for ECI 254 was announced, I started thinking of datasets I could use.
Professor Niemeier had already put Shiny in my mind, since a previous student had used Shiny for their final project.
After some consideration, I remembered my previous work in MAE 253, and my heatmap idea.
I decided that I wanted to try to implement the heatmap in Shiny and see how far I could get.

### Objective

My objective was to create a Shiny application that would display BART passenger data on an actual map.
The application should show where the BART stations and tracks were, as well as color each track by how many passengers traveled on it.
I also wanted users to be able to subset the input data by date and hour, so they could see how many passengers traveled at certain points in time.

### Changes Made After the Presentation

I demonstrated an incomplete version of the application in class on 2019-03-15.
Since then, I have made the following changes (as shown by [my commit history for the repository](https://github.com/jlperona/bart-passenger-heatmap/commits/master)):

* Modify the sidebar to be more useful, and actually use the options that the user input rather than ignoring it.
* Use Shiny's reactives to calculate and update the displayed data based on the user's input. I also coded it so only the tracks are re-rendered on an update, instead of the entire map.
* Validate the user's input and display errors if there's anything wrong. This makes the application more robust. Previously, it would crash if incorrect input was given.

For more detail on the topics listed above, see the Methodology section below.

## Methodology

There are two major portions to this application:

1. Using Python, parsing through the BART data to create an input file.
2. Using Shiny, collating the input file and rendering the heatmap.

### Parsing

The current parsing mechanism is a preparser that only needs to be run once to create the input data for the Shiny application.
The Shiny application then takes that input data, collates it, and displays it.

#### Input Data

BART provides hourly passenger origin-destination data [on their website](https://www.bart.gov/about/reports/ridership).
I found this dataset when I was working on the final project for MAE 253.
At the time of writing, their data ranges from 2011-01-01 to 2018-12-31, which is eight years worth of data.
The files are separated by year, and in their uncompressed form are approximately 250 MB each.

The data takes the following format:

```
2011-05-23,4,FRMT,UCTY,77
```

In order, each column stands for the following:

1. Date, in ISO 8601-style `YYYY-MM-DD` format.
2. Hour, in 24-hour time.
3. Origin station, in BART's 4-letter abbreviation code.
4. Destination station, in BART's 4-letter abbreviation code.
5. The number of passengers who traveled between those two stations.

Combinations of stations for which there were zero passengers that traveled between them are not included.

The BART data can be interpreted as a graph, where the vertices are the stations, edges represent passenger flow, and the weight of each edge is based on how many passengers traveled between that station pair.
The source data, then, can be interpreted as a fully-connected graph, since from any station there can be an entry in the data (and thus an edge in the graph) to any other station.
A fully-connected graph is hard to visualize, however.

#### Prior Work

Instead of attempting to visualize a fully-connected graph, it would be easier if the graph was in the same format as [the BART system map](https://www.bart.gov/system-map).
There would only be edges between adjacent stations on the network.
The edge weights then become how many passengers traveled between those two stations.
Their final destination may not have necessarily been either of those two stations, but they traveled on the track between the two to get to their final destination.

The BART data is not in this format, however.
In my previous work for MAE 253, [`jlperona/bart-hourly-dataset-parser`](https://github.com/jlperona/bart-hourly-dataset-parser), I wrote a parser that parses through the BART origin-destination data and exports a graph file that is in the same format as the system map.
The methodology behind that script is as follows:



#### Parsing On Demand in R

In my Shiny application, I needed the same functionality that the Python script from my previous work gave.
Being able to turn the BART data into something more easily visualized was crucial.

Initially, I tried to replicate my previous Python script using R.
Instead of exporting a graph file, though, I would access the graph's edges 
You can see that attempt in [my repository history](https://github.com/jlperona/bart-passenger-heatmap/blob/a2ccbe602d75cb41e230803973afae65edb0d149/other/igraph.R).
This would mean that the Shiny application would take in the BART data as a direct input.
Since NetworkX doesn't exist in R, I used a package called `igraph` instead.
`igraph` provides network analysis tools, and is available in R.
It provides much of the same functionality that I needed from NetworkX, so it seemed like an appropriate choice.

I was able to successfully implement the parsing using `igraph`, but I came across a problem.
Parsing through a one-day subset of the data with about 26,000 rows would take about 5 minutes to execute in R, which was too long as it was.
Extrapolating this to the entire eight-year dataset would have been infeasible.
Parsing on demand was not the correct solution here.

#### Preparsing in Python

Dijkstra's / BFS

### Shiny Application

#### User Interface (UI)

#### Reactives

#### Validation

## Results

The end result of my work is a working and robust heatmap for the BART data that allows users to subset the input data by date and hour.

### Source Code

I have made all source code for this application available on GitHub in my repository [`jlperona/bart-passenger-heatmap`](https://github.com/jlperona/bart-passenger-heatmap).
The code in that repository is licensed under the MIT License, which is the same license that my prior work [`jlperona/bart-hourly-dataset-parser`](https://github.com/jlperona/bart-hourly-dataset-parser) was licensed under.

I'm a strong believer in the open-source movement.
Making this code available to all under an [open source license](https://opensource.org/licenses) is important to me.
Since I was the only one working on this data and project in the class at the time, I felt comfortable making the code open source before I turned this report in.
That being said, I hope nobody else attempts to pass this work off as their own.

### Running the Application

There are multiple mechanisms by which a user can run the heatmap.
I have listed some of them below.

#### Clone and Run Locally

Users who wish to run the heatmap can download the code and run it locally on their own machine.
One mechanism by which they can do this is via cloning my Git repository using the following command:

```
git clone https://github.com/jlperona/bart-passenger-heatmap.git
```

Once a user has cloned my repository, they can then build the application in RStudio.
This also allows users to make edits to the application, test them out, and potentially build their own similar application.

#### `shiny::runGitHub()`

Another mechanism to run the heatmap locally is provided by a function in the Shiny package.
If the application code is hosted on GitHub, Shiny makes it very easy to launch an application via the [`runGitHub()`](https://www.rdocumentation.org/packages/shiny/versions/0.9.0/topics/runGitHub) function.
It will download and launch Shiny applications that are hosted in a GitHub repository.
If a user wanted to run my application using this function, for instance, the following line of code would do so:

```{r echo=TRUE,eval=FALSE}
shiny::runGitHub("bart-passenger-heatmap", "jlperona")
```

The code above cannot be executed in a RMarkdown document and will give an error if attempted.
It is shown here for informative purposes.

#### *shinyapps.io*

Shiny apps are web apps and thus were meant to be hosted online.
I could have hosted this on a personal website, but I lack a server and the funds to maintain said server.
However, I found out that Shiny apps can be hosted on [*shinyapps.io*](https://www.shinyapps.io/), a website made by the creators of RStudio.

I have hosted my application on *shinyapps.io* at https://jlperona.shinyapps.io/bart-passenger-heatmap/.
This is how I demonstrated the application in my presentation.
However, *shinyapps.io* has tiered pricing depending on usage, limiting the number of hours an application can be run per month.
I am using the free tier, which means that I have a very limited amount of hours.
The link above may not work depending on whether I have exhausted my allotment for a given month.

#### Embedded in RMarkdown

It is possible to embed Shiny applications in RMarkdown, as I've done below.
You can either define the entire application inline using the [`shinyApp()`](https://shiny.rstudio.com/reference/shiny/latest/shinyApp.html) function, or include the application in another directory using the [`shinyAppDir()`](https://shiny.rstudio.com/reference/shiny/latest/shinyApp.html) function.
Since my application is split across multiple files and pulls from other data files, I have opted for the latter function.
In addition, you have to modify the YAML at the top of the document to tell RMarkdown that Shiny content is being included.
At the top of this RMarkdown document, I have added the following line:

```
runtime: shiny
```

You will also need to install some packages to build this RMarkdown document, besides the normal ones typically used to build any RMarkdown document.
The following code should install the necessary packages:

```{r echo=TRUE,eval=FALSE}
install.packages(c("data.table",
                   "fasttime",
                   "leaflet",
                   "rgdal",
                   "shiny")
)
```

The code below is shown for informative purposes.
Unfortunately, the limited width of a RMarkdown document rendered into HTML messes up the UI of my application.
If it is unusable, I'd suggest using one of the methods listed above to run the application instead.
Also, in order to embed the application, I have had to include the entire set of application files with my submission.
The code below will not run unless the application files are included with this RMarkdown document in the path it expects.

```{r echo=TRUE}
# code courtesy of R Markdown: The Definitive Guide by Xie, Allaire, and Grolemund
# directory string cannot be ".."
# https://bookdown.org/yihui/rmarkdown/shiny-embedded.html
shinyAppDir(".",
            options = list(width = "100%",
                           height = 1080)
)
```

## Conclusion

I accomplished the objective I set out to achieve at the beginning of this project: a fully-functional Shiny application that shows a heatmap of the BART system using BART's own data.
In addition, I went slightly further and made the application more robust by validating user input.

On a more personal level, I was able to take an idea I've had in the back of my mind ever since I took MAE 253 and bring it to life.
I also got to learn how to use Shiny to build web apps.
I'm quite happy with the final result and my growth as an R programmer in this class.

### Methods for Improvement

The biggest flaw in this application right now is the UI layout.
I am not a UI or user experience (UX) expert, but I think that the UI can definitely be improved.
However, to my understanding, to do so in Shiny requires Cascading Style Sheets (CSS) experience, which I don't have.

Another "flaw" (which I would consider a difference in opinion) is the layout of the map.
I wanted to see the BART network overlaid on a real map.
However, this means that some of the edges are much larger than the others.
For example, the edges between the San Francisco stations are very difficult to see when the application is initially started.
You have to zoom in in order to see their colors.
A "better" heatmap would make it easier to see all the edges at once, which means that the stations would be evenly spaced.
Due to the setup of this application, it would be relatively easy to create a different heatmap by simply changing the station GIS data.
Since it's open source, anybody could take this as a base and create that application, if they so desire.

### Future Work

There are a couple of ways that I can think of to improve this application at the moment.
All of these should be possible with my skill set.

* Instead of calculating the sum of the number of passengers on an edge, show other summary statistics, such as the mean or median.
* Other visualization types, such as a [chord diagram](https://en.wikipedia.org/wiki/Chord_diagram) that shows passengers' final destinations from a given source station.

I have another idea for future work, which isn't related to improving my application.
I will admit that it was somewhat difficult to get started with Shiny.
[This blog post on DataScience.com](https://www.datascience.com/blog/beginners-guide-to-shiny-and-leaflet-for-interactive-mapping) was the most helpful thing for getting me off the ground, but I was doing some things differently than them, which complicated matters.
That being said, I spent most of my time attempting to figure out a few more advanced pieces of Shiny, such as reactives and validating user input.
In contrast, I was able to get other things, like the preparser, done relatively quickly.
I do think that the Shiny documentation could be better for certain topics, like the ones I've mentioned above.
That is certainly an area for improvement.

## Acknowledgements

I'd like to acknowledge my teammates in MAE 253 who helped me with the previous work that guided this project:

* Baotuan Nguyen, formerly a Master's student in Computer Science at UC Davis. He's currently a software development engineer at Workday, Inc.
* Heidi Schweizer, formerly a PhD student in Agricultural and Resource Economics at UC Davis. She's currently an assistant professor at North Carolina State University.

I'd also like to acknowledge Professor Deb Niemeier, who let me attempt this project even though there was a good chance I'd fail (at least, from my perspective).
I've learned a lot about both R and Shiny through this project, and I also created something which I think is both interesting and useful thanks to her class.
